TBD

### 2019-05-21 manuscript/manuscript.tex
finalized version for resubmission to Am Nat

\begin{center} \large Matthew A. Barbour$^{1,2,\ast}$, Christopher J. Greyson-Gaito$^{2,3}$, Arezoo Sootodeh$^{2}$, Brendan Locke$^{4}$, Jordi Bascompte$^{1}$ \normalsize \end{center}

\noindent 1. University of Zurich, Department of Evolutionary Biology
and Environmental Studies, Zurich, 8057 ZH, Switzerland;

\noindent 2. University of British Columbia, Department of Zoology,
Vancouver, BC V6T 1Z4, Canada;

\noindent 3. University of Guelph, Department of Integrative Biology,
Guelph, ON N1G 2W1, Canada;

\noindent 4. Humboldt State University, Department of Biological
Sciences, Arcata, CA 95521, USA.

\(^\ast\) Corresponding author; e-mail:
\href{mailto:matthew.barbour@ieu.uzh.ch}{\nolinkurl{matthew.barbour@ieu.uzh.ch}}

\bigskip
<!----> \emph{Manuscript elements}: table 1, table 2, figure 1, figure
2, figure 3, figure 4. All figures should be printed in color.

@@

\emph{Number of words}: 3109

@@

Ecological networks, such as food webs describing who-eats-whom in
ecological communities, provide an explicit representation of the
community context \citep{Bascompte2014, McCann2012}. Here, we integrate
ecological networks and adaptive landscapes to understand how the
community context constrains evolutionary change \citep{Arnold1992}.
Different aspects of evolutionary constraints can be inferred by
quantifying the slope and curvature of the adaptive landscape. For
example, the slope is determined by directional selection gradients
acting on each phenotypic trait and influences the trajectory of
evolutionary change \citep{Lande1979, Arnold1992}. Selective constraints
on evolution increase with the number of traits under directional
selection, as this diminishes the number of optimal solutions
\citep{Arnold2003}. The curvature of the adaptive landscape can also
constrain evolution through its indirect effect on genetic constraints
\citep{Arnold1992, Hansen2008}. Genetic constraints are largely governed
by a population's \textbf{G}-matrix ---the additive genetic variances
and covariances between traits \citep{Hansen2008}. In general, genetic
constraints will increase with the number of traits under directional or
stabilizing selection, as this will decrease the additive genetic
variance in those traits \citep{Hansen2008}. Genetic constraints may
also increase with the number of trait combinations under correlational
selection, as this type of selection decreases the evolutionary
independence of traits \citep{Hansen2008}. If we want to predict how the
community context constrains evolutionary change, we must understand how
ecological networks shape the adaptive landscape.

Global change is altering the structure of ecological networks, which
may influence evolutionary constraints in a number of ways. For example,
in a food web, if different consumers impose directional selection on
different traits of a shared resource, then a more diverse consumer
community may constrain evolution by increasing the number of traits
under selection. Alternatively, if consumers impose selection on
different values of a trait, then their selective effects would cancel
each other out in more diverse communities. To examine these different
possibilities, we conducted a field experiment that simulated the
extinction of a consumer guild associated with an abundant insect
herbivore (\emph{Iteomyia salicisverruca}, Family Cecidomyiidae; fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortality from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web structure by either excluding or allowing the guild
of larval parasitoids to impose selection on gall midge and egg
parasitoid traits (fig. \ref{fig:Conceptual}). We then applied modern
statistical methods to quantify how changes in food-web structure
altered the slope and curvature of the adaptive landscape of the gall
midge and egg parasitoid. Taken together, our study gives insight to how
local extinctions may constrain the evolution of interacting
populations.

\begin{figure}
\centering
\includegraphics{analyses/complex_simple_foodwebs_v2.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. To simulate consumer
extinction, we used mesh bags to exclude the guild of larval
parasitoids, only allowing the egg parasitoid (\emph{Platygaster} sp.)
to impose selection. Larval parasitoids include the following species
(from left to right): \emph{Mesopolobus} sp. (Family: Pteromalidae);
\emph{Tetrastichus} sp. (Family: Eulophidae); and \emph{Torymus} sp.
(Family: Torymidae).}
\end{figure}

@@

0.02 [
-0.14,
0.18] \\%& 
%\textbf{
%0.32 [
%0.64,
%0.01] }\\

@@

\begin{figure}
\centering
\includegraphics{analyses/UV_landscapes.pdf}
\caption{\label{fig:UV_Landscape}Adaptive landscape of gall midge
phenotypes in the original food web and with the extinction of larval
parasitoids. Each panel corresponds to a different phenotypic trait:
chamber diameter (A); oviposition preference (B); and clutch size (C).
Bold lines represent selection experienced in the original food web
(orange) and with consumer extinctions (blue). Thin lines represent
bootstrapped replicates to show the uncertainty in selection. For
clarity, we only display 100 bootstraps even though inferences are based
on 1,000 replicates. Note that mean larval survival is plotted on a
natural log scale to reflect the mathematical defintion of the adaptive
landscape.}
\end{figure}

@@

\[\textbf{Curvature}_{\text{Original}} = \begin{pmatrix} 
0 &  &  \\  
0 & 0 &  \\  
0 & 0 & 0.34 \end{pmatrix} - \begin{pmatrix} 
0.34\cdot0.34 &  &  \\  
0\cdot0.34 & 0\cdot0 &  \\  
0\cdot0.34 & 0\cdot0 & 0\cdot0 \end{pmatrix}\]

\[\textbf{Curvature}_{\text{Original}} = \begin{pmatrix} 
-0.12 &  &  \\  
0 & 0 &  \\  
0 & 0 & 0.34 \end{pmatrix}\]

\[\textbf{Curvature}_{\text{Extinction}} = \begin{pmatrix} 
0 &  &  \\  
0 & 0 &  \\  
0 & 0 & 0 \end{pmatrix} - \begin{pmatrix} 
0.21\cdot0.21 &  &  \\  
-0.09\cdot0.21 & -0.09\cdot-0.09 &  \\  
-0.16\cdot0.21 & -0.16\cdot-0.09 & -0.16\cdot-0.16 \end{pmatrix}\]

\[\textbf{Curvature}_{\text{Extinction}} = \begin{pmatrix} 
-0.04 &  &  \\  
0.02 & -0.01 &  \\  
0.03 & -0.01 & -0.03 \end{pmatrix}\]

@@

\begin{figure}
\centering
\includegraphics{analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in the original food web and with
the extinction of larval parasitoids. Each panel corresponds to a
different combination of phenotypic traits: clutch size and chamber
diameter (A); clutch size and oviposition preference (B); oviposition
preference and chamber diameter (C). Arrows represent mean estimates of
directional selection gradients, while contours represent predicted
larval survival of the mean phenotype in each food-web treatment. Notice
that arrows point more toward a corner of the adaptive landscape for
each combination of traits with the extinction of larval parasitoids
compared to the original food web. This indicates that consumer
extinctions more strongly favored trait integration (covariance). Note
that mean larval survival is plotted on a natural log scale to reflect
the mathematical definition of the adaptive landscape.}
\end{figure}

@@

-0.03 [
0.25,
-0.3] \\

@@

0.07 [
0.38,
-0.26] \\

@@

-0.25 [
0.09,
-0.64] \\

@@

-0.05 [
0.33,
-0.43] \\

@@

-0.21 [
0.26,
-0.68] \\

@@

\textbf{
-0.46 [
-0.02,
-1.07] }\\

@@

0 [
0.27,
-0.29] \\

@@

0.25 [
0.6,
-0.05] \\

@@

-0.18 [
0.12,
-0.52] \\ 
\hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Values in brackets represent 95\% confidence intervals. Bold values indicate that the 95\% CI does not overlap zero. $\beta_{\text{Diam}}$ has been adjusted for bias.}
\end{table}

\begin{figure}
\centering
\includegraphics{analyses/selection_on_Platygaster.pdf}
\caption{\label{fig:EggPtoid_Selection}Selection imposed by larval
parasitoids on egg parasitoids (\emph{Platygaster} sp.). The bold line
represents the average difference in the probability of observing the
egg parasitoid (original - extinction of larval parastioids) as a
function of gall midge oviposition preference. Thin lines represent
bootstrapped replicates to show the uncertainty in selection. For
clarity, we only display 100 bootstraps even though inferences are based
on 1,000 replicates. The decrease in the probability of observing egg
parasitoids at high gall-midge densities indicate that larval
parasitoids impose nonlinear selection on egg parasitoids.}
\end{figure}

@@

Our results suggest that the loss of consumers may not only directly
affect connected species, but also result in indirect evolutionary
effects. In our study, this indirect effect arises from egg parasitoids
being released from intraguild predation when we excluded larval
parasitoids. This release occurs more on trees with high larval
densities, which could intensify selection on gall midge oviposition
preference. A growing number of experiments over the past two decades
have demonstrated the presence and potential importance of indirect
evolutionary effects that emerge in a community context
\citep{Pilson1996, Juenger1998, Stinchcombe2001, Lankau2007, Walsh2008, Walsh2010, terHorst2010, Sahli2011, Lau2012, terHorst2015, Schiestl2018, Start2019}.
If indirect evolutionary effects are common \citep{Walsh2013}, then
predicting evolutionary trajectories resulting from consumer extinctions
will require evolutionary studies to explicitly account for the
ecological networks that species are embedded in.

@@

Our study has some limitations to consider when relating our findings to
the effects of consumer extinctions on evolutionary constraints more
generally. First, we removed other sources of mortality prior to our
analyses. While this was necessary to avoid confounding estimates of
selection with effects on phenotypic expression, we may be
overestimating the effects of selection due to parasitoids. Second, our
metrics of evolutionary constraints do not account for variation in the
magnitude of selection pressures. While quantitative metrics are
possible \citep{Hansen2008}, we felt they were not appropriate to apply
here as our experimental removal of larval parasitoids results in a
transient dynamic in the strength (but not direction) of selection. Such
transient dynamics may be common in short-term experimental studies of
natural selection, but they can be detected by using natural history
information and dissecting the sources of selection (e.g.~quantifying
selection surfaces and treatment effects on mean fitness) as we did
here.

>>>>

\emph{Manuscript elements}: table 1, table 2, figure 1, figure 2, figure
3, figure 4. All figures should be printed in color.

@@

\emph{Number of words}: 3108

@@

Ecological networks, such as food webs describing who-eats-whom in
ecological communities, provide an explicit representation of the
community context \citep{Bascompte2014, McCann2012}. Here, we integrate
ecological networks and adaptive landscapes to understand how the
community context constrains evolutionary change. Different aspects of
evolutionary constraints can be inferred by quantifying the slope and
curvature of the adaptive landscape \citep{Arnold1992}. For example, the
slope is determined by directional selection gradients acting on each
phenotypic trait and influences the trajectory of evolutionary change
\citep{Lande1979, Arnold1992}. Selective constraints on evolution
increase with the number of traits under directional selection, as this
diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape can also constrain evolution through
its indirect effect on genetic constraints
\citep{Arnold1992, Hansen2008}. Genetic constraints are largely governed
by a population's \textbf{G}-matrix ---the additive genetic variances
and covariances between traits \citep{Hansen2008}. In general, genetic
constraints will increase with the number of traits under directional or
stabilizing selection, as this will decrease the additive genetic
variance in those traits \citep{Hansen2008}. Genetic constraints may
also increase with the number of trait combinations under correlational
selection, as this type of selection decreases the evolutionary
independence of traits \citep{Hansen2008}. If we want to predict how the
community context constrains phenotypic evolution, we must understand
how ecological networks shape the adaptive landscape.

Global change is altering the structure of ecological networks, which
may influence evolutionary constraints in a number of ways. For example,
in a food web, if different consumers impose directional selection on
different traits of a shared resource, then a more diverse consumer
community may constrain evolution by increasing the number of traits
under selection. Alternatively, if consumers impose selection on
different values of a trait, then their selective effects would cancel
each other out in more diverse communities. To examine these different
possibilities, we conducted a field experiment that simulated the
extinction of a consumer guild associated with an abundant insect
herbivore (\emph{Iteomyia salicisverruca}, Family Cecidomyiidae; fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortality from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web structure by either excluding the guild of larval
parasitoids (extinction food web) or allowing both egg and larval
parasitoids to impose selection on gall midge traits (original food web,
fig. \ref{fig:Conceptual}). We then applied modern statistical methods
to quantify how changes in food-web structure altered the slope and
curvature of the gall midge's adaptive landscape. Taken together, our
study gives insight to how local extinctions may constrain the evolution
of interacting populations.

\begin{figure}
\centering
\includegraphics{../analyses/complex_simple_foodwebs_v3.jpeg}
\caption{\label{fig:Conceptual}Experimental manipulation of food-web
structure associated with a leaf-galling midge (\emph{Iteomyia
salicisverruca}) feeding on the willow \emph{Salix hookeriana}. Black
arrows denote the flow of energy in this network of trophic
interactions. In the original food web, we allowed the full suite of egg
and larval parasitoids to impose selection. To simulate consumer
extinction, we used mesh bags to exclude the guild of larval
parasitoids, only allowing the egg parasitoid (\emph{Platygaster} sp.)
to impose selection. Larval parasitoids include the following species
(from left to right): \emph{Mesopolobus} sp. (Family: Pteromalidae);
\emph{Tetrastichus} sp. (Family: Eulophidae); and \emph{Torymus} sp.
(Family: Torymidae).}
\end{figure}

@@

0.02 [
-0.15,
0.18] \\%& 
%\textbf{
%0.32 [
%0.64,
%0.01] }\\

@@

\begin{figure}
\centering
\includegraphics{../analyses/UV_landscapes.pdf}
\caption{\label{fig:UV_Landscape}Adaptive landscape of gall midge
phenotypes in the original food web and with the extinction of larval
parasitoids. Each panel corresponds to a different phenotypic trait:
chamber diameter (A); oviposition preference (B); and clutch size (C).
Bold lines represent selection experienced in the original food web
(orange) and with consumer extinctions (blue). Thin lines represent
bootstrapped replicates to show the uncertainty in selection. For
clarity, we only display 100 bootstraps even though inferences are based
on 1,000 replicates. Note that mean larval survival is plotted on a
natural log scale to reflect the mathematical defintion of the adaptive
landscape.}
\end{figure}

@@

\[\textbf{Curvature}_{\text{original}} = \begin{pmatrix} 
0 &  &  \\  
0 & 0 &  \\  
0 & 0 & 0.34 \end{pmatrix} - \begin{pmatrix} 
0.34\cdot0.34 &  &  \\  
0\cdot0.34 & 0\cdot0 &  \\  
0\cdot0.34 & 0\cdot0 & 0\cdot0 \end{pmatrix}\]

\[\textbf{Curvature}_{\text{original}} = \begin{pmatrix} 
-0.12 &  &  \\  
0 & 0 &  \\  
0 & 0 & 0.34 \end{pmatrix}\]

\[\textbf{Curvature}_{\text{extinction}} = \begin{pmatrix} 
0 &  &  \\  
0 & 0 &  \\  
0 & 0 & 0 \end{pmatrix} - \begin{pmatrix} 
0.21\cdot0.21 &  &  \\  
-0.09\cdot0.21 & -0.09\cdot-0.09 &  \\  
-0.16\cdot0.21 & -0.16\cdot-0.09 & -0.16\cdot-0.16 \end{pmatrix}\]

\[\textbf{Curvature}_{\text{extinction}} = \begin{pmatrix} 
-0.04 &  &  \\  
0.02 & -0.01 &  \\  
0.03 & -0.01 & -0.03 \end{pmatrix}\]

@@

\begin{figure}
\centering
\includegraphics{../analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in the original food web and with
the extinction of larval parasitoids. Each panel corresponds to a
different combination of phenotypic traits: clutch size and chamber
diameter (A); clutch size and oviposition preference (B); oviposition
preference and chamber diameter (C). Arrows represent mean estimates of
directional selection gradients, while contours represent predicted
larval survival of the mean phenotype in each food-web treatment. Notice
that arrows point more toward a corner of the adaptive landscape for
each combination of traits with the extinction of larval parasitoids
compared to the original food web. This indicates that consumer
extinctions more strongly favored trait integration (covariance). Note
that mean larval survival is plotted on a natural log scale to reflect
the mathematical definition of the adaptive landscape.}
\end{figure}

@@

0.03 [
0.3,
-0.25] \\

@@

-0.07 [
0.26,
-0.39] \\

@@

0.25 [
0.64,
-0.09] \\

@@

0.05 [
0.43,
-0.33] \\

@@

0.21 [
0.68,
-0.26] \\

@@

\textbf{
0.46 [
1.07,
0.02] }\\

@@

0 [
0.29,
-0.27] \\

@@

-0.25 [
0.04,
-0.6] \\

@@

0.18 [
0.52,
-0.12] \\ 
\hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Values in brackets represent 95\% confidence intervals. Bold values indicate that the 95\% CI does not overlap zero. $\beta_{\text{Diam}}$ has been adjusted for bias.}
\end{table}

\begin{figure}
\centering
\includegraphics{../analyses/selection_on_Platygaster.pdf}
\caption{\label{fig:EggPtoid_Selection}Selection imposed by larval
parasitoids on egg parasitoids (\emph{Platygaster} sp.). The bold line
represents the average difference in the probability of observing the
egg parasitoid (original - extinction of larval parastioids) as a
function of gall midge oviposition preference. Thin lines represent
bootstrapped replicates to show the uncertainty in selection. For
clarity, we only display 100 bootstraps even though inferences are based
on 1,000 replicates. The decrease in the probability of observing egg
parasitoids at high gall-midge densities indicate that larval
parasitoids impose nonlinear selection on egg parasitoids.}
\end{figure}

@@

Our results suggest that the loss of consumers may not only directly
affect connected species, but also result in indirect evolutionary
effects. In our study, this indirect effect arises from egg parasitoids
being released from intraguild predation when we excluded larval
parasitoids. This release occurs more on trees with high larval
densities, which could intensify selection on gall midge oviposition
preference. A growing number of experiments over the past two decades
have demonstrated the presence and potential importance of indirect
evolutionary effects that emerge in a community context
\citep{Pilson1996, Juenger1998, Stinchcombe2001, Lankau2007, Walsh2008, Walsh2010, terHorst2010, Sahli2011, Lau2012, terHorst2015, Schiestl2018, Start2019}.
If indirect evolutionary effects are common
\citep{Miller1996, Walsh2013}, then predicting evolutionary trajectories
resulting from consumer extinctions will require evolutionary studies to
explicitly account for the ecological networks that species are embedded
in.

@@

Our study has some limitations to consider when relating our findings to
the effect of consumer extinctions on evolutionary constraints more
generally. First, we removed other sources of mortality prior to our
analyses. While this was necessary to avoid confounding estimates of
selection with effects on phenotypic expression, we may be
overestimating the effects of selection due to consumers. Second, our
metrics of evolutionary constraints do not account for variation in the
magnitude of selection pressures. While metrics that incorporate the
magnitude of selection are possible \citep{Hansen2008}, they were not
appropriate for our study as our experimental removal of larval
parasitoids artificially increased the mean fitness of gall midges (in
the short term), which dampens the magnitude of selection gradients in
this treatment \citep{Hunter2018}. This is why we used more qualitative
metrics of evolutionary constraints, which are robust to this artificial
effect and simply require sufficient sample sizes to detect evidence of
selection.
